---
title: "Statistical Methods for Discrete Response, Time Series, and Panel Data (W271): Lab 4"
author: "Ashesh Choudhury, Kyle Chuang, and Sullivan Swift"
date: "Fall 2018"
output:
  pdf_document: 
    highlight: zenburn
  html_notebook: default
---

```{r setup, include=FALSE}
require(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE,results='asis',warning = FALSE,fig.height=3, collapse = TRUE)
options(xtable.comment=FALSE)
options(xtable.floating=FALSE)
options(xtable.NA.string='NA')
options(xtable.table.placement="!h")
library(Hmisc)
library(car)
library(ggplot2)
library(ggrepel)
library(plm)
library(tidyr)
library(dplyr)
library(xtable)
library(stargazer)
```

```{r include=FALSE}
plottot = function(variable){
  #g=hist(data[variable],main='Histogram of Total Fatality Rate',xlab='Fatalities/population',ylab='Count')
  tmp=data
  tmp['var']=data[,eval(variable)]
  
  #g=hist(data[variable],main=variable,xlab='Fatalities/population',ylab='Count')
  g=ggplot(tmp,aes(x=var))+geom_histogram(fill='blue',color='black',bins=30)+ggtitle(paste('Fatalies/Population','Histogram'))+xlab('Fatalities/Population')+ylab('Count')+theme(plot.title=element_text(hjust=0.5))
  print(g)

  g=ggplot(tmp,aes(y=var,x=factor(year),group=factor(state),color=factor(state)))+geom_line()+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(x=c('Year'),y=c('Fatalities/population'),title=c('Fatalies/Population by Year by State'),color=c('State'))+theme(plot.title = element_text(hjust = 0.5))+guides(color=FALSE)
  print(g)
  
  g=ggplot(tmp,aes(y=var,x=factor(state),fill=factor(state)))+geom_boxplot()+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(x=c('State'),y=c('Fatalities/population'),title=c('Fatalies/Population by State'),fill=c('State'))+theme(plot.title = element_text(hjust = 0.5))+guides(fill=FALSE) 
  print(g)
  
  g=ggplot(tmp,aes(y=var,x=factor(year),fill=factor(year)))+geom_boxplot()+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(x=c('Year'),y=c('Fatalities/population'),title=c('Fatalies/Population by Year'),fill=c('Year'))+theme(plot.title = element_text(hjust = 0.5))+guides(fill=FALSE) 
  print(g)
}

plotmix = function(data,variable,text){
  tmp = data %>% group_by(year) %>% summarize_all(funs(mean))
  tmp['var']=tmp[,eval(variable)]
  g=ggplot(tmp,aes(y=totfatrte,x=year,fill=var))+geom_bar(stat='identity')+labs(x='Year',y='Fatalities/population',title=paste('Fatalities/Population vs ',text,' Laws'))+theme(plot.title = element_text(hjust = 0.5))+guides(fill=guide_legend(title=text))
  print(g)
}

density_plot = function(data, plotvar, title) {
  ggplot(data, aes(plotvar)) + geom_density() + ggtitle(title)
}

conditional_plot = function(data, plotvar, condvar, title) {
  g <- ggplot(data, aes(as.factor(condvar), plotvar)) 
  g + geom_boxplot() + ggtitle(title)
}


ts_resid = function(x){
  par(mfrow=c(2,2),par=c(2,2,2,2))
  plot(x,main='t-Plot')
  qqnorm(x)
  qqline(x,col='red')
  acf(x)
  pacf(x)
  d=data.frame(PhillipsPerron=pp.test(x)$p.value,
               AugmentedDickeyFuller=adf.test(x)$p.value,
               LjungBox=Box.test(x,type='Ljung-Box')$p.value,
               ShapiroWilkes=shapiro.test(x)$p.value)
  xtable(d)
  
  ShapiroWilks <- shapiro.test(x)$p.value
}

plotstackedbox = function(data,target_var,title_text,legend_key,legend_position='bottomleft',ncols=1) {
  tmp=data
  tmp$var=target_var
  tmp$var_bin=ifelse(tmp$var>0,1,0)
  tmp = tmp %>% select(year,totfatrte,var_bin) %>% group_by(year) %>% summarize_all(funs(mean)) %>% mutate(var_none=1-var_bin)
  tmp=tmp$totfatrte*tmp
  tmp$year=tmp$year/sqrt(tmp$totfatrte)
  tmp$totfatrte=NULL
  rownames(tmp)=tmp$year
  tmp$year=NULL
  tmp=t(tmp)
  barplot(as.matrix(tmp),col=c('blue','red'),xlab='Year',ylab='Fatalities/population',main=paste('Fatalities/population vs ',title_text),cex.main=0.7,cex.axis=0.7,cex.lab=0.7,cex.names=0.7)
  legend(legend_position,legend=c(legend_key,paste('No ',legend_key)),col=c('blue','red'),lty=c(1,1,1,1),cex=0.7,bg='white',ncol=ncols)
}

plotstackedbox_sl = function() {
  tmp=data %>% select(year,totfatrte,sl55,sl65,sl70,sl75,slnone) %>% group_by(year) %>% summarize_all(funs(mean)) 
  tmp=tmp$totfatrte*tmp
  tmp$year=tmp$year/sqrt(tmp$totfatrte)
  tmp$totfatrte=NULL
  rownames(tmp)=tmp$year
  tmp$year=NULL
  tmp=t(tmp)
  barplot(as.matrix(tmp),col=c('red','blue','grey','yellow','black'),xlab='Year',ylab='Fatalities/population',main='Fatalities/population by year and Speed Limit Mix',cex.main=0.7,cex.lab=0.7,cex.axis=0.7,cex.names=0.7)
  legend('bottomleft',legend=rownames(tmp),col=c('red','blue','grey','yellow','black'),lty=c(1,1,1,1),cex=0.4,bg='white')
}

plotstackedbox_bac = function() {
  tmp = data %>% select(year,totfatrte,bac10,bac08) %>% group_by(year) %>% summarize_all(funs(mean)) %>% mutate(bac_none=1-bac10-bac08) 

  tmp=tmp$totfatrte*tmp
  tmp$year=tmp$year/sqrt(tmp$totfatrte)
  tmp$totfatrte=NULL
  rownames(tmp)=tmp$year
  tmp$year=NULL
  tmp=t(tmp)
  barplot(as.matrix(tmp),col=c('blue','green','red'),xlab='Year',ylab='Fatalities/100k vehicle miles',main='Fatalities/100k vehicle mile by year and BAC Law Mix',cex.main=0.7,cex.lab=0.7,cex.axis=0.7,cex.names=0.7)
  legend('topright',legend=rownames(tmp),col=c('blue','green','red'),lty=c(1,1,1,1),cex=0.7,bg='white',ncol=3)
}


plotstackedbox_seatbelt = function() {
  tmp = data %>% select(year,totfatrte,sbprim,sbsecon) %>% group_by(year) %>% summarize_all(funs(mean)) %>% mutate(sbnone=1-sbprim-sbsecon) 
  
  tmp=tmp$totfatrte*tmp
  tmp$year=tmp$year/sqrt(tmp$totfatrte)
  tmp$totfatrte=NULL
  rownames(tmp)=tmp$year
  tmp$year=NULL
  tmp=t(tmp)
  barplot(as.matrix(tmp),col=c('blue','green','yellow'),xlab='Year',ylab='Fatalities/100k vehicle miles',main='Fatalities/100k vehicle mile by year and Seatbelt Laws',cex.main=0.7,cex.lab=0.7,cex.axis=0.7,cex.names=0.7)
  legend('topright',legend=rownames(tmp),col=c('blue','green','yellow'),lty=c(1,1,1,1),cex=0.7,bg='white',ncol=3)
}
```

```{r include=FALSE}
#rm(list=ls())
```

```{r}
load('driving.RData')
```

*Note: Custom functions have been used to create this pdf. They are not included in the R pdf but are in the R-markdown file.*

(Question 1) The dataset contains about 1200 observations ranging from 1980 to 2004 for the 48 continental states. Speed limits (`slXX`) seat belt and zero tolerance, graduated driver, blood alcohol level (`bacXX`), per se law variables are in percent of year by months in from $0$ to $1$. `Sbl70plus`, `sbprim`, `sbsecon` and `dXX` variables are simply derivatives or dummy variables of the other variables in the data set. Other variables are continuous with a a base of 0 and no top coding, except perc14_24 (100%).

Our research question is whether or not traffic laws affect total fatalities. Total fatalities is a function of population, vehicle miles, traffic laws and unobservable variables. Our dataset contains 9 fatality-related variables; some normalized in various ways. We will not consider the weekend and night fatality variables as we are focused on total fatalities and not when they occurred.

(Question 1 & 2) Three states, 5, 44 and 10, significantly increased in population while other states were relatively flat. This suggests using a population-normalized fatality measure, such as `totfatrte`, for examining traffic laws' impact on driving fatalities.

```{r}
ggplot(data,aes(y=statepop,x=factor(year),group=factor(state),color=factor(state)))+geom_line()+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(x=c('Year'),y=c('State Population'),title=c('State Population by Year'),color=c('State'))+theme(plot.title = element_text(hjust = 0.5)) +guides(color=FALSE)
```

(Question 1 & 2) Vehicle miles has roughly similar trends for almost all states. This further confirms `totfatrte` as a dependent variable for traffic law causal inference as vehicle miles is likely to be correlated with changes in population. It is also relatively "stable" from year-on-year (YoY) change .

```{r}
ggplot(data,aes(y=vehicmiles,x=factor(year),group=factor(state),color=factor(state)))+geom_line()+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(x=c('Year'),y=c('Vehicle Miles'),title=c('Vehicle Miles by Year by State'),color=c('State'))+theme(plot.title = element_text(hjust = 0.5))+ guides(color=FALSE)
```

(Question 1 & 2) Mean fatalities/population has dropped by about $10\%$ from 1980 to 2004. State 51 has persistently stayed near the top of the fatalies while State 38 seems to stay near the bottom. The range of fatalities changes throughout time, with the highest fatality rate dropping from over $50$ in 1980 to under $45$ in 2004. The minimum values for fatalities drop as well, but less overall, from about just over $10$ in 1980 to just under $10$ in 2004.

There are several states that have higher overall fatality rates than the others consistently - states 25, 32 and 51 may be further explored to see the relationships with the observed variables.

```{r}
plottot('totfatrte')
```

(Question 1) Initially, the speed limit (`slXX`) increase does not seem to affect the average total fatality rate across states immediately. In 1986-1987, speed limits increased in many states. In 1988-1991, fatalities fell by about 10%. The variable may be a candidate as an interaction variable with time. 

```{r}
plotstackedbox_sl()
```

To further examine speed limits, we split the variables into two groups - states with speed limits under 70 and states with speed limits over 70. From the chart below, it appears that speed limit over 70+ do not impact fatalities in a significant way.

```{r}
plotstackedbox(data,data$sl70plus,'Speed Limit > 70','Speed Limit > 70 Mix')
```

`bac_none` represents states that do not have laws relating to blood alcohol content.

Blood Alcohol Content laws appear to have an immediate impact on fatalities. More interestingly, the graph suggests transforming the BAC laws into a binary variable, as  `bac08` and `bac10` do not appear to affect fatalities. The initial implementation of drinking-related laws do an impact. 

```{r include=FALSE}
#t.test(data[data$year==2001 & data$bac10==1,"totfatrte"],data[data$year==2001 & data$bac08==1,'totfatrte'])
#t.test(data[data$year==2002 & data$bac10==1,"totfatrte"],data[data$year==2002 & data$bac08==1,'totfatrte'])  
```

We performed cross-sectional t-tests for 2001 and 2002 when the `bac08` and `bac10` were closest to 50% between the states. The two t-tests are performed cross-sectionally to avoid the general downward trend of fatalities impacting the analysis. In both t-tests, we fail to reject the null hypothesis, $H_0$: differences in means $=0$. From the graph below, there aren't likely to be lagged effects for `bac10` and `bac08` as `bac10` decreased from 1997-2004, but fatalities appear roughly the same. The lack of lagged effect makes sense as bac laws will immediately impact drunk driver and remove them from the roads.

We may later test the joint probability of `bac08` and `bac10`. If both show insignificance in a multivariate regression but rejects $H_0$ jointly, we should convert it into a binary variable of BAC laws or none.

```{r}
plotstackedbox_bac()
```

Seatbelts appear to have a simultanenous decrease with fatalities. It should be included as an independent variable. Much like BAC levels, the mix of seatbelt laws does not appear to affect fatalities. There appears to be some lagged effect on the binary seatbelt law - possibly the population is getting in the habit of putting on seatbelts. After most states have implemented at least primary seatbelt laws (1992-1995), the average fatality rate flattens.

We performed cross-sectional t-tests for 1999 and 2000 where the percentages are closer to even for `sbprim` and `sbsecon`. For both t-tests, we fail to reject $H_0$ and there does not appear to be any contemporaneous differences between seatbelt laws. Despite the decrease of `sbsecon` mix from 1995 to 2004, the fatalities remained flat. There is unlikely to be a lagged impact of seatbelt law differences. Finally, we also note that State 30 does not have seatbelt laws throughout the period if further analysis on specifics of the seatbelt law is required.

```{r}
plotstackedbox_seatbelt()
```

```{r include=FALSE}
#t.test(data[data$year==1999 & data$sbprim==1,"totfatrte"],data[data$year==1999 & data$sbsecon==1,'totfatrte'])
#t.test(data[data$year==2000 & data$sbprim==1,"totfatrte"],data[data$year==2000 & data$sbsecon==1,'totfatrte'])  
```

Minimum drinking age appears to have some effect on fatalies. The mean `minage` from 1980-1990 trended higher to $21$ in 1990. 

```{r}
plotmix(data,'minage','Minimum Drinking Age')
```

Interestingly, all states listed below do not have minimum age laws change during the period. If we were to focus on `minage`, we can split the data into 2 sets and run separate analysis, detrend and analyze the impact of `minage` on fatalities. `minage` could be interacted this variable with BAC variables as raising minimum drinking age may potentially offset some effect of BAC laws. The interaction term is expected to have a negative coefficient.

```{r}
tmp2=data %>% group_by(state) %>% summarize_all(funs(mean)) %>% as.data.frame
tmp2=tmp2[tmp2$minage %in% unique(data$minage),c('state','minage')]
tmp2=tmp2[!(tmp2$state %in% c(47,51)),c('state','minage')]
xtable(t(tmp2[1:7,]), table.placement="!h")
xtable(t(tmp2[8:dim(tmp2)[1],]), table.placement="!h")
```

Zero tolerance laws do not appear to have a contemperous impact on fatalities based on the changes in laws from 1992-1997. It may potentally have a long-tailed effect.

```{r}
plotstackedbox(data,data$zerotol,'Zero Tolerance Law','Zero Tol. ','topright',2)
```

Graduated driver license laws do not appear to impact fatalies very much. Most likely, it will not impact fatalities even accounting for time lags. The changes from graduated license laws from 1999 to 2004 barely impacted fatalities with or without lag effects.

```{r}
plotstackedbox(data,data$gdl,'Gradulated Drivers Law','GDL')
```

Per se law does not appear to impact fatalities. The increase from 1982 to 1983 did not appear to have a contemporaneous or lagged impact on fatalities. Per se laws may have interactions with BAC laws as it increases the "harshness" of BAC laws.

```{r}
plotstackedbox(data,data$perse,'Per Se Law','Per Se Law','topright',2)
```

Vehicle miles per capita may be secondarily affected by traffic laws, such as graduated license laws. Preliminarily, fatalities appear to decrease as it increases. This relationship is counter-intuitive and is likely to be a trending effect through time. Other independent variables may have a stronger impact than the number of miles driven per capita on total fatalities rates.

```{r}
plotmix(data,'vehicmilespc','Vehicle Miles Per Capita')
```

While all states increased `vehicmilespc`, state 46 was interesting in that there was a large increase in 2001-2001 followed by a large drop back down to historical trend by 2003-2004. Closer, state-specific reasoning may be required. Given that we do not have that information, we will not examine it further.

```{r}
ggplot(data,aes(y=vehicmilespc,x=factor(year),group=factor(state),color=factor(state)))+geom_line()+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(x=c('Year'),y=c('Vehicle Miles/Capita'),title=c('Vehicle Miles/Capita by Year by State'),color=c('State'))+theme(plot.title = element_text(hjust = 0.5))+guides(color=FALSE)
```

Fatalities do not appear to be affected by unemployment rates. In fact, the pattern appears somewhat random. We expect this variable to have a $\beta_{unem}$ close to 0 in regressions. If we do include the variable, we may want to consider logging it to spread out the values and improve the regression. It does not appear heteroskedastic.

```{r}
plotmix(data,'unem','Unemployment')
```

Percent of peopl ages 14-24 is left skewed and is transformed with a log function. 
```{r}
par(mfrow=c(1,2))
hist(data$perc14_24,main='Historgram of Percent 14-24',xlab='Percent 14-24',ylab='Count',cex.main=0.7,cex.axis=0.7,cex.lab=0.7,cex.name=0.7)
hist(log(data$perc14_24),main='Historgram of Log(Percent 14-24)',xlab='Log(Percent 14-24)',ylab='Count',cex.main=0.7,cex.axis=0.7,cex.lab=0.7,cex.name=0.7)
```

It appears that percent 14-24 is negatively correlated with fatalities. This may potentially be an artifact of the general trend in both population and fatalities.

```{r}
tmp=data
tmp$logperc14_24=log(data$perc14_24)
tmp = data %>% group_by(year) %>% mutate(logperc=log(perc14_24)) %>% summarize_all(funs(mean))
ggplot(tmp,aes(y=totfatrte,x=year,fill=logperc))+geom_bar(stat='identity')+labs(x='Year',y='Fatalities/population',title='Fatalities/Population vs Percent 14-24')#+theme(plot.title = element_text(hjust = 0.5))+guides(fill=guide_legend(title=text))
```

While all states decreased in their population of 14-24 year olds, a few states increased in the ratio. Most significantly, state 45's increase stood out amongst all the states with the sudden jump and decrease between 2001-2003. Again, without more state specific information, it's difficult to further examine it.

```{r}
ggplot(data,aes(y=perc14_24,x=factor(year),group=factor(state),color=factor(state)))+geom_line()+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(x=c('Year'),y=c('Percent 14-24'),title=c('Percent 14-24 by States Over Time'),color=c('State'))+theme(plot.title = element_text(hjust = 0.5))+guides(color=FALSE)
```

Due to limited scope, bivariate EDA will focus on variables that we have identified as potential interesting interaction variables - `bac`, `perse` and `minage`. We will binarize BAC laws to streamline the analysis. BAC + Per Se Law seems to have a time lagged effect on decreasing total fatalities as seen from 1984-1992.

```{r}
tmp=data
tmp$bac=ifelse(tmp$bac08>=1 | tmp$bac10>=1,1,0)
tmp$gdlperse=tmp$gdl*tmp$perse
tmp$bacperse=tmp$bac*tmp$perse
tmp$minageperse=tmp$minage*tmp$perse
tmp$bacminage=tmp$bac*tmp$minage
plotstackedbox(tmp,tmp$bacperse,'BAC + Per Se Laws','BAC + Per Se Laws','topright',2)
```

Minimum age with Per Se Laws does not appear to affect fatalities and is unlikely to be a significant interaction variable contemporaneously or lagged.

```{r}
plotstackedbox(tmp,tmp$minageperse,'Minimum Age + Per Se Laws','Minimum Age+ Per Se Laws','topright',2)
```

Again, the BAC + Minimum Age Law relationship appears to have a time lagged component much like BAC + Per Se Laws. It appears to have a negative correlation as expected. However, this may simply be because of the BAC laws. T-tests and time-lagged variables may be done to further the analysis.

```{r}
plotstackedbox(tmp,tmp$bacperse,'BAC + Minimum Age Laws','BAC + Minium Age Laws','bottomright',2)
```

From the EDA, there are variables that appear to be "incorrectly" correlated with fatalities, such as `vehicmilespc`. Others such as BAC may be better transformed into a binary on-off variable. Per se, zero tolerance, graduated license, minimum drinking age laws appears to have no effect while seatbelts appear to have a contemporaneous impact on fatalites. Speed Limit laws appear to have a lagged effect.

(Question 2) We will first examine the general time trend of fatalities. Recall that our fatalities variable, `totfatrte`, is fatalities per 100,000 population is already  normalized by population, so proper analysis of impact of traffic laws on fatalities can be analyzed. Note that traffic laws are most likely uncorrelated with state population as shown in the beginning of the report. Generally, states laws have lower correlation with population suggesting that it has less impact than vehicle miles, which may be affected by state laws. Therefore, total fatalities adjusted by population is likely a better variable to infer state law causalities as noted in the beginning of the report.

```{r}
xtable(cor(data[,c('statepop','vehicmiles','minage','zerotol','gdl','seatbelt')]))
```

The means of `totfatrte` is through time is below. 

```{r}
tmp=data %>% select(year,totfatrte) %>% group_by(year) %>% summarize_all(funs(mean)) %>% as.data.frame
xtable(t(tmp[1:9,]),caption=c('Mean Fatalities/Population'),caption.placement='top')
xtable(t(tmp[10:18,]))
xtable(t(tmp[19:25,]))
```

A simple plot of the mean `totfatrte` shows that it has decreased over time.

```{r}
ggplot(data %>% select(year,totfatrte) %>% group_by(year) %>% summarize_all(funs(mean)),aes(year,totfatrte,label=totfatrte))+geom_line(color='red')+geom_point(color='blue')+geom_text(aes(label=round(totfatrte,1)),color='black',hjust=0, vjust=-1)+ggtitle('Mean Fatalities/Population across all States by Year')+theme(plot.title = element_text(hjust = 0.5))
tmp = data %>% select(year,totfatrte) %>% group_by(year) %>% summarise_all(funs(mean))
```
```{r include=FALSE}
ggplot(tmp,aes(year,totfatrte,label=totfatrte))+geom_line()+geom_point(aes(col='red'))+geom_label_repel(aes(label = round(totfatrte,2)), box.padding   = 0.35,  point.padding = 0.5, segment.color = 'grey50') + theme_classic()
```

(Question 2) Regression of fatalities vs each year shows that there is a clear significant downward trend. The $F-statistic$ of 7.16 is a p-value of $\sim 0$ shows that the dummy variables for year is jointly significant. We only show the first few coefficients below and the rest are plotted with the standard error. The regression suggests that fatalities have been decreasing through time and the $\beta$s show the mean differential between the year $t$ and 1980. The intercept of the regression is the mean fatalities in 1980 and the coefficients is the mean differences from 1980 for each year respectively. Notice the 2 chart (above & below) are exact same shape after the 1st year (1980). Driving became much safer over time. Note that it's a general trend over time as the +/-95% confidence intervals of $\beta$s overlap each other in the surrounding years. We can only conclude that the trend has decreased but it is difficult to firmly draw conclusions on any given year. 

```{r}
m=lm(totfatrte~factor(year),data)
xtable(summary(m)$coefficients[1:5,])
```
```{r}
summary(m)$fstatistic
```
```{r}
tmp=data.frame(confint(m))
tmp=tmp[2:dim(tmp)[1],]
colnames(tmp)=c('lower','upper')
tmp$year=1981:2004
tmp$beta=(tmp$lower+tmp$upper)/2
ggplot(tmp,aes(x=year,y=beta))+geom_point(size=2.5,color='red')+geom_line()+geom_errorbar(aes(ymax=tmp$upper,ymin=tmp$lower),color='blue') +labs('title'='Coefficients for Fatality/Population by Year (Mean difference from 1980)',x='Year',y="Fatality difference from 1980")+geom_point(aes(col='red'))+geom_label_repel(aes(label = round(tmp$beta,2)), box.padding   = 0.15,  point.padding = 0.5, segment.color = 'grey50')+guides(color=FALSE)+theme(plot.title=element_text(hjust=0.5))#+geom_point(color='red')+geom_text(aes(label=round(tmp$beta,1)),color='black',hjust=0, vjust=1.5)
```

Note that the residuals are not normally distributed and fails the Shapiro Wilks test. They are also serially correlated as shown by Durbin-Watson test, indicating potentially inefficiency.

```{r results='markup'}
shapiro.test(m$residuals)
durbinWatsonTest(m)
```

(Question 3) We will now expand the previous regression with additional regressors - `bac08`, `bac10`, `perse`, `sbprim`, `sbsecon`, `sl70plus`, `gld`, `perc14_24`, `unem` and `vehicmilespc`. `perc14-24` is logged since it is very left skewed and `unem` is also logged. Logging the values expands the distribution, thus leading to a better regression results. `vehicmilepc` does not appear to require transformations as it appears more normally/uniformly distributed. The rest of variables are binary and no transformations are performed. BAC and speed limit variables are not binarized and no interactions are implemented due to the limited scope of the analysis.

```{r}
library(plm)
data.p=pdata.frame(data,index = c('state','year'))
m=lm(totfatrte~factor(year)+bac08+bac10+perse+sbprim+sbsecon+sl70plus+gdl+log(perc14_24)+log(unem)+vehicmilespc,data)
xtable(summary(m)$coefficients[26:35,])
```

We did not not show all the $\beta$s for the factor year but they can be found in the R code. `bac08` and `bac10` have coefficients of `r round(data.frame(t(m$coefficients))$bac08,3)` and `r round(data.frame(t(m$coefficients))$bac10,3)` respectively. 

Holding all other variables constant,

* The $\beta_{bac08}$ and $\beta_{bac10}$ represent the impact of having `bac08` and `bac10` laws in that year on the fatalities across the states, with $p<.01$.
* Per se laws are also associated with a decrease fatalities by `r round(data.frame(t(m$coefficients))$perse,3)` once enacted, with $p<.01$.
* Primary seat belt laws appears to correspond with a decrease the total fatalities rate with $\beta_{sbprim}=$ `r round(data.frame(t(m$coefficients))$sbprim,3)` that is marginally significant , $p<.1$. 

We note that all the $\beta$ estimators, p-values and other regression statistics above are biased, inconsistent and inefficient as fixed effects and explanatory variable are likely to be correlated. In reality, no conclusion or inference can be made with the estimators since the regression assumptions are violated and estimators are biased. We note that the residuals fail both the Shapiro-Wilks normality test and Durbin-Watson serial correlation test.

```{r results='markup'}
shapiro.test(m$residuals)
durbinWatsonTest(m)
```

Additionally, we can do a pool test to check if the coefficients are consistent vs a fixed effects model. The null hypothesis, $H_0$, is rejected showing that the estimates are not consistent - showing that the estimators are biased.

```{r results='markup'}
pooltest(totfatrte~bac08+bac10+perse+sbprim+sbsecon+sl70plus+gdl+log(perc14_24)+log(unem)+vehicmilespc,data = data.p,model='within')
```

(Question 4) Given the pooled model assumptions may be violated, we will examine the data using a fixed effects model.

```{r}
m.fe=plm(totfatrte~bac08+bac10+perse+sbprim+sbsecon+sl70plus+gdl+log(perc14_24)+log(unem)+vehicmilespc,data = data.p,model='within')
stargazer(m.fe,no.space = T,header=FALSE,table.placement='!h',float=FALSE)
```

We note that the residuals are not normal and serially correlated.

```{r results='markup'}
shapiro.test(m.fe$residuals)
pbgtest(m.fe,order=2)
```

```{r}
d=data.frame(pooled_coef=summary(m)$coefficients[26:35,1],
             pooled_SE=summary(m)$coefficients[26:35,2],
             FE_coef=summary(m.fe)$coefficients[,1],
             FE_SE=summary(m.fe)$coefficients[,2]) 
xtable(d)
```

(Question 4) The coefficients are different between the pooled OLS and Fixed Effects regression. The downward bias from pooled OLS due to fixed effects appears in `bac08` and `log(perc14_24)`. `bac10` and `vehicmilespc` are roughly unbiased. `perse`, `gds`, `sbprim`, and `sbsecon` have been biased upwards. More importantly, `sl70plus` and `log(unem)` have both been biased by fixed effects such that the signs of the coefficients are incorrect. No inferences can really be made from the regression as the standard errors are incorrect due to serially correlated residuals.

FE model is better than pooled OLS since it removes the fixed effects. Pooled OLS assumes  no correlation between unobserved variables and any of the regressors. It is highly likely this assumption is violated. There are many unobserved variables that could be correlated with our regressors. For example, dry laws, which are unobserved, may be correlated with `bac08` laws and affect fatalities. From the pool test and above, we can tell the estimators are biased in the pooled OLS since the estimators are incorrect. Therefore, correlation of fixed effects and explanatory variables definitely exist. The bias causing incorrect signs of coefficients in `sl70plus` and `log(unem)` is notable. For the FE models, the assumption is that the idiosyncratic errors ($\epsilon_{it}$) are uncorrelated conditional on the indepedendent variables and time-invariant unobservable variables. Given the current context, the FE assumptions are more reasonable as time-invariant variables can be eliminated. Though FE errors are still serially correlated preventing us from doing proper inference, we will at least have unbiased estimators. In the case of pooled OLS, the estimators will not be unbiased and inference cannot be done either.

(Question 5) In comparing FE models with RE models, FE models is likely to be a better estimate in the current context. Like the Pooled OLS, RE model assumes no correlation between fixed effects and explanatory variables. The difference between the pooled OLS and RE models is that RE corrects the serial correlation within the composite error by estimating a correlation. The advantage of RE models over FE is the ability to estimate time-invariant variables. However, it also requires an extremely strong assumption that fixed effects are independent of all explanatory variables across all time periods. Given the endogeneity issues with pooled OLS shown above, we believe fixed effects is a much better model than random effects for the current data set. We can run a Hausman test where the $H_0$ assumes the errors are not correlated with the regressors. The $H_0$ is rejected in the Hausman test; therefore, a fixed effects model is more appropriate for the current dataset.

```{r results='markup'}
m.re=plm(totfatrte~bac08+bac10+perse+sbprim+sbsecon+sl70plus+gdl+log(perc14_24)+log(unem)+vehicmilespc,data = data.p,model='random')
phtest(m.fe,m.re)
```

```{r include=FALSE}
summary(m.fe)
```

The final model is
\[ \begin{aligned}
y_{it}-\bar{y}_i=& -1.84 (x_{(bac08)it}-\bar{x}_{i(bac08)}) +  -1.47 (x_{(bac10)it}-\bar{x}_{i(bac10)}) -1.67 (x_{(per se)it}-\bar{x}_{i(per se)}) \\
&+ -1.73 (x_{(sbprim)it}-\bar{x}_{i(sbprim)}) +-0.78 (x_{(sbsecon)it}-\bar{x}_{i(sbsecon)}) \\
&+ -1.16 (x_{(sl70plus)it}-\bar{x}_{i(sl70plus)})  + -0.61 (x_{(gdl)it}-\bar{x}_{i(gdl)} \\ 
&+ 14.54(x_{(log(per14_24))it}-\bar{x}_{i(log(perc14_24))})+-3.30(x_{(log(unem))it}-\bar{x}_{i(log(unem))})\\
&+0.0003(x_{(vehicmilespc)it}-\bar{x}_{i(vehicmilespc)})+(\epsilon_{it}-\bar{\epsilon}_{i})
\end{aligned}
\]

```{r}
tmp=data %>% group_by(state) %>% summarize_all(funs(mean)) %>% select(state,vehicmilespc) %>% as.data.frame
l=NULL
for (i in tmp[,1]){
  l=c(l,sd(data[data$state==i,'vehicmilespc']-tmp[tmp$state==i,'vehicmilespc']))
}
```

(Question 6) If $vehicmilepc$ increase by 1,000 in a time period $t$ assuming the mean, $\bar{vehicmilepc}$, does not change in the FE model, $totfatrte$ is expected to increase `r round(data.frame(t(m.fe$coefficients))$vehicmilespc,3)` with a 95% confidence interval of `r round(data.frame(t(confint(m.fe)))$vehicmilespc,3)`. The $\beta_{vehicmilespc}$ may not be an extremely precise estimate. We note that the standard deviation after differencing the mean on vehicmilespc is `r round(sd(l),3)` while the standard deviation without differencing is `r round(sd(data$vehicmilespc),3)`. The smaller variation across time causes the estimate to be less precise.

(Question 7) Assuming we have no omitted variable bias (FE model) with autocorrelation and heteroskedasticity existing in the errors, the estimators will be inefficient but unbiased. The standard errors estimated on the estimators are incorrect as the variance formula used for estimation is incorrect. It is inefficient as it is no longer the minimum variance estimator. The estimators, however, remain unbiased.